---
title: "decipher_loughran_scrna"
output: html_document
date: "2026-03-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  engine.path = list(bash = "/bin/bash")
)

knitr::opts_knit$set(
  root.dir = "/standard/CookLab/Ryan/plasmidsaurus_de"
)
```

```{r}
library(tidyverse)
library(data.table)
library(dplyr)
library(stringr)
library(ggplot2)
library(functions)
library(DESeq2)
library(vsn)
library(pheatmap)
library(openxlsx)
library(Biostrings)
library(stringdist)
library(Seurat)

loughran_seurat <- readRDS("20260312_loughran_scrna.rds")


```

First, I just checked the file structure. It's somewhat long but worth scrolling through initially
```{bash, echo = TRUE}
source /etc/profile.d/modules.sh
module load tree
tree /standard/CookLab/loughran_scrna/fastq
```

It looks like jeff sorted samples and/or libraries into directories. I make sure each directory has only one prefix, which would inidicate to me they are one demultiplexed sample and/or library across lanes. As of now I am unsure the difference between novaseq1 directory and novaseq2 directory but I check them both. The flow cell ID seems to differ so I suppose 2 different runs?
```{bash}
for d in /standard/CookLab/loughran_scrna/fastq/novaseq1/outs/fastq_path/H5CFNDSXY/*_GEX; do
  echo "== $d =="
  ls "$d" | sed -E 's/_S[0-9]+_L00[0-9]_(I1|R1|R2)_001.fastq.gz$//' | sort | uniq -c
done
for d in /standard/CookLab/loughran_scrna/fastq/novaseq2/outs/fastq_path/H5MVCDSXY/*_GEX; do
  echo "== $d =="
  ls "$d" | sed -E 's/_S[0-9]+_L00[0-9]_(I1|R1|R2)_001.fastq.gz$//' | sort | uniq -c
done
```
I see jeff2 in both novaseq1 and novaseq2 but how can I check if they are the same sample? Also, there seem to be more samples in the Reports directory

```{bash}
(echo "novaseq1"; ls /standard/CookLab/loughran_scrna/fastq/novaseq1/Reports/html/H5CFNDSXY/UVA_LoughranLab/) | paste - <(echo "novaseq2"; ls /standard/CookLab/loughran_scrna/fastq/novaseq2/Reports/html/H5MVCDSXY/UVA_LoughranLab/) | column -t

```

Never mind, I think these are just reports, not fastq files. Maybe some of this are other projects?
```{bash}
source /etc/profile.d/modules.sh
module load tree
tree /standard/CookLab/loughran_scrna/fastq/novaseq1/Reports/html/H5CFNDSXY/UVA_LoughranLab/jeff2_GEX_S13
```

Back to the novaseq1 and novaseq2 fastq files. I think these are just two runs of the same samples 1-8
```{bash}
(echo "novaseq1"; ls /standard/CookLab/loughran_scrna/fastq/novaseq1/outs/fastq_path/H5CFNDSXY/) | paste - <(echo "novaseq2"; ls /standard/CookLab/loughran_scrna/fastq/novaseq2/outs/fastq_path/H5MVCDSXY/) | column -t

```




Next, I think I should check for number of unique values in the metadata columns. Ideally something matches 8. Or perhaps 7 since sample 1 seems to be missing GEX
```{r}
values <-  data.frame(
    column = names(loughran_seurat@meta.data),
    unique_values = sapply(loughran_seurat@meta.data, function(col) length(unique(col))),
    row.names = NULL
  )
values[values$unique_values == 7 | values$unique_values == 8, ]
unique(loughran_seurat$batch)
unique(loughran_seurat$chain_pairing)
```


Let's look at barcodes
```{r}
loughran_metadata <- loughran_seurat@meta.data
loughran_metadata <- cbind(functions::split_column(data.frame("barcodes" = rownames(loughran_metadata)), "barcodes", "-"), loughran_metadata)
unique(loughran_metadata$barcodes_2)
unique(loughran_metadata$barcodes_3)
```

Bingo! 7 samples, probably corresponding to jeff# that have GEX. I don't know what/who xing is
```{r}
sample_from_barcode <- str_extract(colnames(loughran_seurat), "(?<=-)[1-7](?=-xing5p$)")
names(sample_from_barcode) <- colnames(loughran_seurat)

loughran_seurat <- AddMetaData(loughran_seurat,
                               sample_from_barcode,
                               col.name = "sample_from_barcode")
loughran_metadata <- loughran_seurat@meta.data
seur_obj_list <- SplitObject(loughran_seurat, split.by = "sample_from_barcode")

seur_one <- seur_obj_list[[1]]
seur_two <- seur_obj_list[[2]]
seur_three <- seur_obj_list[[3]]
seur_four <- seur_obj_list[[4]]
seur_five <- seur_obj_list[[5]]
seur_six <- seur_obj_list[[6]]
seur_seven <- seur_obj_list[[7]]

meta_one <- seur_one@meta.data
unique(meta_one$group)
meta_two <- seur_two@meta.data
unique(meta_two$group)
meta_three <- seur_three@meta.data
unique(meta_three$group)
meta_four <- seur_four@meta.data
unique(meta_four$group)
meta_five <- seur_five@meta.data
unique(meta_five$group)
meta_six <- seur_six@meta.data
unique(meta_six$group)
meta_seven <- seur_seven@meta.data
unique(meta_seven$group)

unique(loughran_metadata$group)
```


Okay I think the experimental structure is something like this

7 biological samples (jeff1–jeff7)
   ×
3 library types (GEX, FB, TCR)
   ×
4 sample indexes each
   ×
2 NovaSeq runs
   ×
4 lanes per run

```{bash}
#pwd
#cd /standard/CookLab/loughran_scrna/fastq/novaseq1/outs/fastq_path/H5CFNDSXY/jeff2_GEX
#pwd
#ls
#zcat *_R1_*.fastq.gz | sed -n '2~4p' | cut -c1-16 > cell_barcodes.txt
#zcat *_I1_*.fastq.gz | sed -n '2~4p' > sample_indexes.txt
#paste sample_indexes.txt cell_barcodes.txt > index_cell_pairs.txt
```

Based on the data below, I believe it may be possible to demultiplex the 4 pooled samples by their i5 indices. Each lane of a given library has only 4 i5 barcodes and they are the same 4 across lanes. Additionally, novseq1 appear to be the same as novaseq2. Next I should just align the fastq within each jeff# directory and worry about trying to demultiplex after I see if there are hits or not
```{r}
pair_barcode_indices <- fread("/standard/CookLab/loughran_scrna/fastq/novaseq1/outs/fastq_path/H5CFNDSXY/jeff2_GEX/index_cell_pairs.txt", header = FALSE)

novaseq1_lane_barcode <- fread("/standard/CookLab/loughran_scrna/fastq/novaseq1_laneBarcode.tsv", sep = "\t", header = TRUE)
novaseq1_lane_barcode <- split_column(novaseq1_lane_barcode, "sample", "_")
novaseq1_lane_barcode$dir <- paste0(novaseq1_lane_barcode$sample_1, "_", novaseq1_lane_barcode$sample_2)
jeff2_GEX <- dplyr::filter(novaseq1_lane_barcode, `dir` == "jeff2_GEX")
jeff3_GEX <- dplyr::filter(novaseq1_lane_barcode, `dir` == "jeff3_GEX")
View(cbind(jeff2_GEX$barcode_sequence, jeff3_GEX$barcode_sequence))

novaseq2_lane_barcode <- fread("/standard/CookLab/loughran_scrna/fastq/novaseq2_laneBarcode.tsv", sep = "\t", header = TRUE)
novaseq2_lane_barcode <- split_column(novaseq2_lane_barcode, "sample", "_")
novaseq2_lane_barcode$dir <- paste0(novaseq2_lane_barcode$sample_1, "_", novaseq2_lane_barcode$sample_2)
jeff2_GEX_2 <- dplyr::filter(novaseq2_lane_barcode, `dir` == "jeff2_GEX")
jeff3_GEX_2 <- dplyr::filter(novaseq2_lane_barcode, `dir` == "jeff3_GEX")
View(cbind(jeff2_GEX_2$barcode_sequence, jeff3_GEX_2$barcode_sequence))

View(cbind(jeff2_GEX$barcode_sequence, jeff2_GEX_2$barcode_sequence, jeff3_GEX$barcode_sequence, jeff3_GEX_2$barcode_sequence))
cbind(jeff2_GEX$barcode_sequence, jeff2_GEX_2$barcode_sequence, jeff3_GEX$barcode_sequence, jeff3_GEX_2$barcode_sequence)
```


As of now, I am not sure which seurat barcode -# relates to which jeff# fastq. For now I will assume -1 is jeff2 as both suffixes are the lowest number

Regardless, it doesn't seem like the i5 indices are specific to any particular group in the seurat metadata. For T_2074_untreated_PBMC barcodes, it looks like there are all 4 i5 barcodes

This chunk does confirm jeff2 = -1, jeff3 = -2, etc

```{r}
colnames(pair_barcode_indices) #first time I read in this enormous file I didn't hae header = FALSE so this will need to change when the time comes
T_2074_one <- gsub("-1-xing5p", "", row.names(meta_one[which(meta_one$group == "T_2074_untreated_PBMC"),]))
test_one <- pair_barcode_indices[which(pair_barcode_indices$V2 %in% T_2074_one == TRUE),]

nrow(test)/nrow(pair_barcode_indices)
```

```{r}
check_10x_barcodes_in_fastq <- function(fastq_file, barcodes, barcode_length = 16) {
  # Basic checks
  if (!file.exists(fastq_file)) {
    stop("FASTQ file does not exist: ", fastq_file)
  }
  
  if (length(barcodes) == 0) {
    stop("You must provide at least one barcode.")
  }
  
  # Clean barcode list
  barcodes <- unique(toupper(trimws(barcodes)))
  barcodes <- barcodes[nzchar(barcodes)]
  
  if (length(barcodes) == 0) {
    stop("No valid barcodes remain after cleaning.")
  }
  
  # Open FASTQ (supports .gz or plain text)
  con <- if (grepl("\\.gz$", fastq_file, ignore.case = TRUE)) {
    gzfile(fastq_file, open = "rt")
  } else {
    file(fastq_file, open = "r")
  }
  
  on.exit(close(con))
  
  found <- setNames(rep(FALSE, length(barcodes)), barcodes)
  
  repeat {
    lines <- readLines(con, n = 400000)
    if (length(lines) == 0) break
    
    # FASTQ sequence lines are every 2nd line of each 4-line record
    seq_lines <- lines[seq(2, length(lines), by = 4)]
    
    # Extract first barcode_length bases
    observed_barcodes <- substr(toupper(seq_lines), 1, barcode_length)
    
    # Mark matches
    matched <- intersect(barcodes, observed_barcodes)
    if (length(matched) > 0) {
      found[matched] <- TRUE
    }
    
    # Stop early if all barcodes have already been found
    if (all(found)) break
  }
  
  n_found <- sum(found)
  n_total <- length(barcodes)
  percent_found <- 100 * n_found / n_total
  
  list(
    n_barcodes_input = n_total,
    n_barcodes_found = n_found,
    percent_found = percent_found,
    found_barcodes = names(found)[found],
    missing_barcodes = names(found)[!found]
  )
}
cancer_celltypes <- c("TLGLL", "NKLGLL")
normal_celltypes <- setdiff(unique(loughran_seurat@meta.data$celltype), cancer_celltypes)
jeff2_cancer_cell <- list(gsub("-1-xing5p", "", row.names(meta_one[which(meta_one$celltype %in% cancer_celltypes),])), "jeff2", "S4",
                   "cancer_celltypes")
jeff2_normal_cell <- list(gsub("-1-xing5p", "", row.names(meta_one[which(meta_one$celltype %in% normal_celltypes),])), "jeff2", "S4",
                   "normal_celltypes")


jeff3_cancer_cell <- list(gsub("-2-xing5p", "", row.names(meta_two[which(meta_two$celltype %in% cancer_celltypes),])), "jeff3", "S5",
                   "cancer_celltypes")
jeff3_normal_cell <- list(gsub("-2-xing5p", "", row.names(meta_two[which(meta_two$celltype %in% normal_celltypes),])), "jeff3", "S5",
                   "normal_celltypes")



jeff4_cancer_cell <- list(gsub("-3-xing5p", "", row.names(meta_three[which(meta_three$celltype %in% cancer_celltypes),])), "jeff4", "S6",
                     "cancer_celltypes")
jeff4_normal_cell <- list(gsub("-3-xing5p", "", row.names(meta_three[which(meta_three$celltype %in% normal_celltypes),])), "jeff4", "S6",
                     "normal_celltypes")



jeff5_cancer_cell <- list(gsub("-4-xing5p", "", row.names(meta_four[which(meta_four$celltype %in% cancer_celltypes),])), "jeff5", "S15",
                     "cancer_celltypes")
jeff5_normal_cell <- list(gsub("-4-xing5p", "", row.names(meta_four[which(meta_four$celltype %in% normal_celltypes),])), "jeff5", "S15",
                     "normal_celltypes")


jeff6_cancer_cell <- list(gsub("-5-xing5p", "", row.names(meta_five[which(meta_five$celltype %in% cancer_celltypes),])), "jeff6", "S16",
                     "cancer_celltypes")
jeff6_normal_cell <- list(gsub("-5-xing5p", "", row.names(meta_five[which(meta_five$celltype %in% normal_celltypes),])), "jeff6", "S16",
                     "normal_celltypes")


jeff7_cancer_cell <- list(gsub("-6-xing5p", "", row.names(meta_six[which(meta_six$celltype %in% cancer_celltypes),])), "jeff7", "S17",
                   "cancer_celltypes")
jeff7_normal_cell <- list(gsub("-6-xing5p", "", row.names(meta_six[which(meta_six$celltype %in% normal_celltypes),])), "jeff7", "S17",
                   "normal_celltypes")



jeff8_cancer_cell <- list(gsub("-7-xing5p", "", row.names(meta_seven[which(meta_seven$celltype %in% cancer_celltypes),])), "jeff8", "S18",
                          "cancer_celltypes")
jeff8_normal_cell <- list(gsub("-7-xing5p", "", row.names(meta_seven[which(meta_seven$celltype %in% normal_celltypes),])), "jeff8", "S18",
                          "normal_celltypes")


```

```{r}
subset_10x_fastq_by_barcodes <- function(
  r1_fastq_gz,
  r2_fastq_gz,
  barcodes,
  out_r1_gz,
  out_r2_gz,
  barcode_length = 16,
  chunk_size = 100000
) {
  # Basic checks
  if (!file.exists(r1_fastq_gz)) stop("R1 file does not exist: ", r1_fastq_gz)
  if (!file.exists(r2_fastq_gz)) stop("R2 file does not exist: ", r2_fastq_gz)
  if (length(barcodes) == 0) stop("You must provide at least one barcode.")
  if (chunk_size <= 0) stop("chunk_size must be > 0")
  
  # Clean barcode list
  barcodes <- unique(toupper(trimws(barcodes)))
  barcodes <- barcodes[nzchar(barcodes)]
  if (length(barcodes) == 0) stop("No valid barcodes remain after cleaning.")
  
  # Open connections
  r1_con <- gzfile(r1_fastq_gz, open = "rt")
  r2_con <- gzfile(r2_fastq_gz, open = "rt")
  out1_con <- gzfile(out_r1_gz, open = "wt")
  out2_con <- gzfile(out_r2_gz, open = "wt")
  
  on.exit({
    close(r1_con)
    close(r2_con)
    close(out1_con)
    close(out2_con)
  }, add = TRUE)
  
  total_reads <- 0L
  kept_reads <- 0L
  
  repeat {
    # Read chunks as FASTQ records (4 lines per read)
    r1_lines <- readLines(r1_con, n = chunk_size * 4)
    r2_lines <- readLines(r2_con, n = chunk_size * 4)
    
    if (length(r1_lines) == 0 && length(r2_lines) == 0) break
    
    if (length(r1_lines) != length(r2_lines)) {
      stop("R1 and R2 differ in number of lines within a chunk. Files may be unsynchronized.")
    }
    
    if (length(r1_lines) %% 4 != 0) {
      stop("R1 chunk does not contain complete FASTQ records.")
    }
    
    if (length(r2_lines) %% 4 != 0) {
      stop("R2 chunk does not contain complete FASTQ records.")
    }
    
    n_reads <- length(r1_lines) / 4L
    total_reads <- total_reads + n_reads
    
    # Sequence lines are line 2 of each 4-line FASTQ record
    r1_seq_idx <- seq(2, length(r1_lines), by = 4)
    r1_seqs <- r1_lines[r1_seq_idx]
    
    # Extract putative 10x barcode from start of R1 sequence
    observed_barcodes <- substr(toupper(r1_seqs), 1, barcode_length)
    keep <- observed_barcodes %in% barcodes
    
    if (any(keep)) {
      kept_reads <- kept_reads + sum(keep)
      
      # Expand logical vector from reads to FASTQ lines
      keep_lines <- rep(keep, each = 4)
      
      writeLines(r1_lines[keep_lines], out1_con, sep = "\n")
      writeLines(r2_lines[keep_lines], out2_con, sep = "\n")
    }
  }
  
  list(
    total_reads_processed = total_reads,
    reads_kept = kept_reads,
    percent_reads_kept = 100 * kept_reads / total_reads,
    output_r1 = out_r1_gz,
    output_r2 = out_r2_gz
  )
}

barcode_lists <- ls(envir = .GlobalEnv)[sapply(ls(envir = .GlobalEnv), function(x) inherits(get(x, envir = .GlobalEnv), "list"))]
barcode_lists <- barcode_lists[-which(barcode_lists == "seur_obj_list")]
if(length(barcode_lists) != 14){
  stop()
}

log_file <- "/standard/CookLab/loughran_scrna/uncompressed_fastq/test_demultiplex.log"
zz <- file(log_file, open = "wt")
sink(zz, split = TRUE)
sink(zz, type = "message")
#first dir is home directory for some reason
#testing on just _1_750000 and _750001_whatever. When ready, fix the for loop initiation
# for(i_dir in list.dirs("/standard/CookLab/loughran_scrna/uncompressed_fastq/")[2:length(list.dirs("/standard/CookLab/loughran_scrna/uncompressed_fastq/"))]){
for(i_dir in list.dirs("/standard/CookLab/loughran_scrna/uncompressed_fastq/")[2:length(list.dirs("/standard/CookLab/loughran_scrna/uncompressed_fastq/"))][1:2]){
  print(paste0("Starting ", i_dir, " ", Sys.time()))
  for(i_file in list.files(i_dir, pattern = ".*R2.*fastq.gz$", full.names = TRUE)){
    print(paste0("Starting ", i_file, " ", Sys.time()))
    temp_r2 <- i_file
    temp_r1 <- gsub("_R2_", "_R1_", temp_r2)
    for(i_barcode in barcode_lists[grep(substr(basename(temp_r1), 1, 5), barcode_lists)]){
      temp_barcodes <- get(i_barcode)[[1]]
      if(grepl("cancer", i_barcode)){
        i_celltype <- "cancer"
      }
      if(grepl("normal", i_barcode)){
        i_celltype <- "normal"
      }
      temp_r1_out <- gsub("fastq.gz", paste0(i_celltype,".out.fastq.gz.celltype.demultiplex"), temp_r1)
      temp_r2_out <- gsub("fastq.gz", paste0(i_celltype,".out.fastq.gz.celltype.demultiplex"), temp_r2)
      subset_10x_fastq_by_barcodes(temp_r1,
                                    temp_r2,
                                    temp_barcodes,
                                    temp_r1_out,
                                    temp_r2_out,
                                    barcode_length = 16,
                                    chunk_size = 100000)
    }
  }
  print(paste0("Finished ", i_dir, " ", Sys.time()))
}
sink(type = "message")
sink()
close(zz)

# for(i_barcode in barcode_lists){
#   temp_obj <- get(i_barcode)
#   temp_barcodes <- temp_obj[[1]]
#   temp_dir <- temp_obj[[2]]
#   temp_s <- temp_obj[[3]]
#   for(i_lane in c("L001", "L002", "L003", "L004")){
#     temp_ns1_r1 <- paste0("/standard/CookLab/loughran_scrna/fastq/subset_reads_R1/novaseq1.subset.",
#                         temp_dir,
#                         "_GEX_",
#                         temp_s,
#                         "_",
#                         i_lane,
#                         "_R1_001.fastq.gz")
#     temp_ns1_r2 <- paste0("/standard/CookLab/loughran_scrna/fastq/subset_reads/novaseq1.subset.",
#                           temp_dir,
#                           "_GEX_",
#                           temp_s,
#                           "_",
#                           i_lane,
#                           "_R2_001.fastq.gz")
#     temp_ns1_r1_out <- paste0("/standard/CookLab/loughran_scrna/fastq/subset_reads_custom_demultiplex_celltype/",
#                         temp_obj[[4]],
#                         "_novaseq1.subset.",
#                         temp_dir,
#                         "_GEX_",
#                         temp_s,
#                         "_",
#                         i_lane,
#                         "_R1_001.fastq.gz")
#      temp_ns1_r2_out <- paste0("/standard/CookLab/loughran_scrna/fastq/subset_reads_custom_demultiplex_celltype/",
#                         temp_obj[[4]],
#                         "_novaseq1.subset.",
#                         temp_dir,
#                         "_GEX_",
#                         temp_s,
#                         "_",
#                         i_lane,
#                         "_R2_001.fastq.gz")
#      subset_10x_fastq_by_barcodes(temp_ns1_r1,
#                                   temp_ns1_r2,
#                                   temp_barcodes,
#                                   temp_ns1_r1_out,
#                                   temp_ns1_r2_out,
#                                   barcode_length = 16,
#                                   chunk_size = 100000)
#      temp_ns2_r1 <- paste0("/standard/CookLab/loughran_scrna/fastq/subset_reads_R1/novaseq2.subset.",
#                         temp_dir,
#                         "_GEX_",
#                         temp_s,
#                         "_",
#                         i_lane,
#                         "_R1_001.fastq.gz")
#     temp_ns2_r2 <- paste0("/standard/CookLab/loughran_scrna/fastq/subset_reads/novaseq2.subset.",
#                           temp_dir,
#                           "_GEX_",
#                           temp_s,
#                           "_",
#                           i_lane,
#                           "_R2_001.fastq.gz")
#     temp_ns2_r1_out <- paste0("/standard/CookLab/loughran_scrna/fastq/subset_reads_custom_demultiplex_celltype/",
#                         temp_obj[[4]],
#                         "_novaseq2.subset.",
#                         temp_dir,
#                         "_GEX_",
#                         temp_s,
#                         "_",
#                         i_lane,
#                         "_R1_001.fastq.gz")
#      temp_ns2_r2_out <- paste0("/standard/CookLab/loughran_scrna/fastq/subset_reads_custom_demultiplex_celltype/",
#                         temp_obj[[4]],
#                         "_novaseq2.subset.",
#                         temp_dir,
#                         "_GEX_",
#                         temp_s,
#                         "_",
#                         i_lane,
#                         "_R2_001.fastq.gz")
#      subset_10x_fastq_by_barcodes(temp_ns2_r1,
#                                   temp_ns2_r2,
#                                   temp_barcodes,
#                                   temp_ns2_r1_out,
#                                   temp_ns2_r2_out,
#                                   barcode_length = 16,
#                                   chunk_size = 100000)
#      rm(temp_ns1_r1, temp_ns1_r2, temp_ns1_r1_out, temp_ns1_r2_out)
#   }
#   rm(temp_obj, temp_barcodes, temp_dir, temp_s)
#   
# }
# t_2074 <- 
# subset_10x_fastq_by_barcodes(
#   "/standard/CookLab/loughran_scrna/fastq/subset_reads_R1/novaseq1.subset.jeff2_GEX_S4_L001_R1_001.fastq.gz",
#   "/standard/CookLab/loughran_scrna/fastq/subset_reads/novaseq1.subset.jeff2_GEX_S4_L001_R2_001.fastq.gz",
#   T_2074_one,
#   "/standard/CookLab/loughran_scrna/fastq/subset_reads_R1/T_2074_untreated_PBMC_novaseq1.subset.jeff2_GEX_S4_L001_R1_001.fastq.gz",
#   "/standard/CookLab/loughran_scrna/fastq/subset_reads/T_2074_untreated_PBMC_novaseq1.subset.jeff2_GEX_S4_L001_R2_001.fastq.gz",
#   barcode_length = 16,
#   chunk_size = 100000
# )
```
